---
title: "Lithium Battery Materials: Band Gap Prediction"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
set.seed(42)
```

## Libraries

```{r libraries}
library(tidyverse)
library(MASS)
library(rpart)
library(rpart.plot)
library(tidymodels)   # parsnip, dials, tune, rsample, workflows
library(ranger)       # RF engine for tidymodels
library(xgboost)      # XGBoost engine for tidymodels
library(pROC)         # AUC-ROC
```

## Data Loading & Preprocessing

Note: run `augment_columns.py` in the `chem-calorine` conda environment first to generate
`lithium_battery_materials_augmented.csv`.

```{r load}
df <- read_csv("lithium_battery_materials_augmented.csv")

drop_cols <- c("material_id", "formula_pretty", "elements", "band_gap")

df_clean <- df %>%
  dplyr::select(-all_of(drop_cols)) %>%
  mutate(
    is_semiconductor          = factor(is_semiconductor, levels = c(FALSE, TRUE),
                                       labels = c("No", "Yes")),
    contains_transition_metal = as.integer(contains_transition_metal)
  ) %>%
  drop_na() %>%
  rename_with(make.names)   # "MagpieData mean X" -> "MagpieData.mean.X"

cat("Dimensions:", nrow(df_clean), "x", ncol(df_clean), "\n")
table(df_clean$is_semiconductor)
```

## 80 / 10 / 10 Train / Val / Test Split

```{r split}
n         <- nrow(df_clean)
train_idx <- sample(seq_len(n), size = floor(0.8 * n))
remaining <- setdiff(seq_len(n), train_idx)
val_idx   <- sample(remaining, size = floor(0.5 * length(remaining)))
test_idx  <- setdiff(remaining, val_idx)

train <- df_clean[train_idx, ]
val   <- df_clean[val_idx,   ]
test  <- df_clean[test_idx,  ]

train_val <- rbind(train, val)   # used by RF (tidymodels) and XGBoost (caret)

y_val  <- val$is_semiconductor
y_test <- test$is_semiconductor

cat(sprintf("Train: %d | Val: %d | Test: %d\n", nrow(train), nrow(val), nrow(test)))
```

## Helper: Evaluation Metrics

```{r eval_fn}
evaluate <- function(actual, predicted_class, predicted_prob, label = "") {
  if (nchar(label) > 0) cat(label, "\n")
  acc     <- mean(predicted_class == actual)
  roc_obj <- roc(actual, predicted_prob, levels = c("No", "Yes"), quiet = TRUE)
  auc_val <- as.numeric(auc(roc_obj))
  cat(sprintf("  Accuracy : %.4f\n", acc))
  cat(sprintf("  AUC-ROC  : %.4f\n", auc_val))
  print(table(Predicted = predicted_class, Actual = actual))
  invisible(list(acc = acc, auc = auc_val, roc = roc_obj))
}
```

---

## 1. Logistic Regression

Train a **full model** and a **stepwise-selected model** (bidirectional AIC) on the training
set, compare on the validation set, then report test metrics for the winner.

```{r logistic_train}
make_lr_df <- function(df) {
  df %>%
    mutate(y = as.integer(is_semiconductor == "Yes")) %>%
    dplyr::select(-is_semiconductor)
}
lr_train <- make_lr_df(train)
lr_val   <- make_lr_df(val)
lr_test  <- make_lr_df(test)

lr_full <- glm(y ~ ., data = lr_train, family = binomial())
lr_step <- MASS::stepAIC(lr_full, direction = "both", trace = FALSE)

cat("Full model — features    :", length(coef(lr_full)) - 1, "\n")
cat("Stepwise model — features:", length(coef(lr_step)) - 1, "\n")
cat("Stepwise retained:\n")
print(names(coef(lr_step))[-1])
```

```{r logistic_val}
to_class <- function(prob) factor(ifelse(prob > 0.5, "Yes", "No"), levels = c("No", "Yes"))

full_prob_val <- predict(lr_full, newdata = lr_val, type = "response")
step_prob_val <- predict(lr_step, newdata = lr_val, type = "response")

full_val_res <- evaluate(y_val, to_class(full_prob_val), full_prob_val, "Full LR (val)")
step_val_res <- evaluate(y_val, to_class(step_prob_val), step_prob_val, "Stepwise LR (val)")

lr_best      <- if (full_val_res$auc >= step_val_res$auc) lr_full else lr_step
lr_best_name <- if (full_val_res$auc >= step_val_res$auc) "Full LR" else "Stepwise LR"
cat("Selected:", lr_best_name, "\n")
```

```{r logistic_test}
# NOTE: not retraining on validation and test here
lr_prob_test <- predict(lr_best, newdata = lr_test, type = "response")
lr_res <- evaluate(y_test, to_class(lr_prob_test), lr_prob_test,
                   paste(lr_best_name, "(test)"))
```

---

## 2. Decision Tree

Pruned using rpart's internal 10-fold CV with the **1-SE rule**: pick the simplest tree
whose `xerror` is within one standard error of the minimum. The validation set is used
for evaluation only (consistent with the other models).

```{r decision_tree}
dt_model <- rpart(
  is_semiconductor ~ .,
  data    = train,
  method  = "class",
  control = rpart.control(cp = 1e-4, xval = 10)
)

# 1-SE rule
cp_tbl    <- dt_model$cptable
min_row   <- which.min(cp_tbl[, "xerror"])
threshold <- cp_tbl[min_row, "xerror"] + cp_tbl[min_row, "xstd"]
best_cp   <- cp_tbl[max(which(cp_tbl[, "xerror"] <= threshold)), "CP"]
dt_pruned <- prune(dt_model, cp = best_cp)

rpart.plot(dt_pruned, type = 4, extra = 104, main = "Decision Tree (1-SE pruned)")

dt_prob_val   <- predict(dt_pruned, newdata = val,  type = "prob")[, "Yes"]
dt_class_val  <- predict(dt_pruned, newdata = val,  type = "class")
dt_prob_test  <- predict(dt_pruned, newdata = test, type = "prob")[, "Yes"]
dt_class_test <- predict(dt_pruned, newdata = test, type = "class")

evaluate(y_val,  dt_class_val,  dt_prob_val,  "Decision Tree (val)")
dt_res <- evaluate(y_test, dt_class_test, dt_prob_test, "Decision Tree (test)")
```

---

## 3. Random Forest (tidymodels + Bayesian optimisation)

Tune `mtry` (1-10), `min_n` (1-100), and `max.depth` (3-50) using Bayesian optimisation
against the explicit validation set (no CV). `tune_bayes()` starts with 5 random
configurations then runs 10 further iterations guided by a Gaussian process surrogate,
for 15 total evaluations.

```{r random_forest}
# Build a single-split resample representing our explicit val set
val_split <- rsample::make_splits(
  list(analysis   = seq_len(nrow(train)),
       assessment = seq_len(nrow(val)) + nrow(train)),
  data = train_val
)
val_rset <- rsample::manual_rset(list(val_split), ids = "Validation")

rf_rec  <- recipe(is_semiconductor ~ ., data = train_val)

# NOTE: trees should also be tuned

rf_spec <- rand_forest(
  mtry  = tune(),
  min_n = tune(),
  trees = 500
) %>%
  set_engine("ranger", max.depth = tune(), importance = "impurity") %>%
  set_mode("classification")

rf_wf <- workflow() %>%
  add_recipe(rf_rec) %>%
  add_model(rf_spec)

# Define parameter ranges; max.depth has no built-in dials entry so supply one
max_depth_param <- new_quant_param(
  type      = "integer",
  range     = c(3L, 50L),
  label     = c(max.depth = "Max Tree Depth"),
  inclusive = c(TRUE, TRUE)
)

rf_params <- rf_wf %>%
  extract_parameter_set_dials() %>%
  update(
    mtry      = mtry(range  = c(1L, 10L)),
    min_n     = min_n(range = c(1L, 100L)),
    max.depth = max_depth_param
  )

rf_bayes <- tune_bayes(
  rf_wf,
  resamples  = val_rset,
  param_info = rf_params,
  iter       = 10,          # 5 initial + 10 BO = 15 total
  initial    = 5,
  metrics    = metric_set(roc_auc),
  control    = control_bayes(verbose = TRUE, no_improve = 10)
)

autoplot(rf_bayes, type = "performance") +
  ggtitle("RF Bayesian Optimisation: AUC-ROC per Iteration")

cat("Best RF hyperparameters:\n")
print(select_best(rf_bayes, metric = "roc_auc"))
```

```{r rf_test}
rf_best_params <- select_best(rf_bayes, metric = "roc_auc")
rf_final_wf    <- finalize_workflow(rf_wf, rf_best_params)
rf_final_fit   <- fit(rf_final_wf, data = train_val)

rf_prob_test  <- predict(rf_final_fit, new_data = test, type = "prob")$.pred_Yes
rf_class_test <- predict(rf_final_fit, new_data = test)$.pred_class
rf_res <- evaluate(y_test, rf_class_test, rf_prob_test, "Random Forest (test)")

# Variable importance from the fitted ranger model
rf_imp <- rf_final_fit %>%
  extract_fit_engine() %>%
  `[[`("variable.importance")
rf_imp <- sort(rf_imp, decreasing = TRUE)
barplot(rf_imp[1:10], las = 2, main = "RF Variable Importance (top 10)", cex.names = 0.7)

# Save a train-only model (no val data) for downstream regression scripts.
# Best hyperparameters are still those found via val-set Bayesian optimisation.
rf_train_only_fit <- fit(rf_final_wf, data = train)
saveRDS(rf_train_only_fit, "classifier.rds")
cat("Saved classifier.rds\n")
cat("Load in another script with:\n")
cat("  rf_model <- readRDS('classifier.rds')\n")
cat("  predict(rf_model, new_data = df, type = 'prob')$.pred_Yes\n")
cat("  predict(rf_model, new_data = df)$.pred_class\n")
```

---

## 4. Gradient Boosted Trees (tidymodels + Bayesian optimisation)

Tune `trees` (100-500), `tree_depth` (3-10), and `learn_rate` (0.01-0.32); other
parameters fixed at sensible defaults. Reuses `val_rset` from section 3.

```{r xgboost}
xgb_spec <- boost_tree(
  trees      = tune(),
  tree_depth = tune(),
  learn_rate = tune()
) %>%
  set_engine("xgboost",
             colsample_bytree = 1,
             verbosity        = 0) %>%
  set_mode("classification")

xgb_wf <- workflow() %>%
  add_recipe(recipe(is_semiconductor ~ ., data = train_val)) %>%
  add_model(xgb_spec)

xgb_params <- xgb_wf %>%
  extract_parameter_set_dials() %>%
  update(
    trees      = trees(range      = c(100L, 5000L)),
    tree_depth = tree_depth(range = c(3L, 50L)),
    learn_rate = learn_rate(range = c(-2, log10(0.32)))  # 0.01–0.32, log10 scale
  )

xgb_bayes <- tune_bayes(
  xgb_wf,
  resamples  = val_rset,   # same single-split rset as RF
  param_info = xgb_params,
  iter       = 10,         # 5 initial + 10 BO = 15 total
  initial    = 5,
  metrics    = metric_set(roc_auc),
  control    = control_bayes(verbose = TRUE, no_improve = 10)
)

autoplot(xgb_bayes, type = "performance") +
  ggtitle("XGBoost Bayesian Optimisation: AUC-ROC per Iteration")

cat("Best XGBoost hyperparameters:\n")
print(select_best(xgb_bayes, metric = "roc_auc"))
```

```{r xgb_test}
xgb_best_params <- select_best(xgb_bayes, metric = "roc_auc")
xgb_final_wf    <- finalize_workflow(xgb_wf, xgb_best_params)
xgb_final_fit   <- fit(xgb_final_wf, data = train_val)

xgb_prob_test  <- predict(xgb_final_fit, new_data = test, type = "prob")$.pred_Yes
xgb_class_test <- predict(xgb_final_fit, new_data = test)$.pred_class
xgb_res <- evaluate(y_test, xgb_class_test, xgb_prob_test, "XGBoost (test)")
```

---

## Final Model Comparison (Test Set)

```{r comparison}
results <- tibble(
  Model    = c(lr_best_name, "Decision Tree", "Random Forest", "XGBoost"),
  Accuracy = c(lr_res$acc,   dt_res$acc,      rf_res$acc,      xgb_res$acc),
  AUC_ROC  = c(lr_res$auc,   dt_res$auc,      rf_res$auc,      xgb_res$auc)
)
print(results %>% arrange(desc(AUC_ROC)))

plot(lr_res$roc,  col = "blue",   main = "ROC Curves (Test Set)", lwd = 2)
plot(dt_res$roc,  col = "orange", add = TRUE, lwd = 2)
plot(rf_res$roc,  col = "green",  add = TRUE, lwd = 2)
plot(xgb_res$roc, col = "red",    add = TRUE, lwd = 2)
legend("bottomright",
  legend = c(lr_best_name, "Decision Tree", "Random Forest", "XGBoost"),
  col    = c("blue", "orange", "green", "red"),
  lwd    = 2)
```
